---
title: "Earthdata API Practice with HTTR"
format: 
  html:
    self-contained: true
editor: visual
---

There exists [an online paper trail](https://urs.earthdata.nasa.gov/documentation/for_users/data_access/r) detailing an old deprecated workflow using `httr` to hit the earthdata api and return amazon s3 information to access the data.

The original code is below:

```{r}
#| label: httr script
#| eval: false

# Set up R
# You may need to install the httr package.
# install.packages("httr")
library(httr)

netrc_path <- "/path/to/.netrc"
cookie_path <- "/path/to/.urs_cookies"
downloaded_file_path <- "/path/to/filename"

# Before using the script
#Set up your ~/.netrc file as listed here: https://wiki.earthdata.nasa.gov/display/EL/How+To+Access+Data+With+cURL+And+Wget
set_config(config(followlocation=1,netrc=1,netrc_file=netrc_path,cookie=cookie_path,cookiefile=cookie_path,cookiejar=cookie_path))
httr::GET(url = "https://disc2.gesdisc.eosdis.nasa.gov/data/TRMM_RT/TRMM_3B42RT_Daily.7/2000/03/3B42RT_Daily.20000301.7.nc4",
                  write_disk(downloaded_file_path, overwrite = TRUE))
```

This code could likely be updated using the `httr2` package to provide similar functionality to `earthaccess`. This is not something I am skilled in so I did not get far.

It also appears that the `httr` package was used within the [earthdatalogin package here](https://github.com/boettiger-lab/earthdatalogin/blob/81c9d7f521c415254534f13f6a1b4f6dc7e4a514/R/edl_search.R#L5) for the api access. Might be an oppurtunity to take advantage of the new package (if I understood how any of this was working).

### Earthdata API

Details on the earthdata api can be found [here](https://cmr.earthdata.nasa.gov/search/site/docs/search/api.html#general-request-details)

```{r}
#| label: edl-search as httr2

# This is the httr function used:
# https://github.com/boettiger-lab/earthdatalogin/blob/81c9d7f521c415254534f13f6a1b4f6dc7e4a514/R/edl_search.R#L5


#install.packages("httr2")
library(httr2)

# # Another S3 package that might be worth using:
# install.packages("aws.s3")

# This isn't targeted enough:
# earthdata_granules_url <- "https://cmr.earthdata.nasa.gov/search/granules" # Need to be more targeted
# "errors":["The CMR does not allow querying across granules in all collections...

# What/how do I find the api url?
earthdata_api <- "https://obdaac-tea.earthdatacloud.nasa.gov/ob-cumulus-prod-public"
eosdis_url <- "https://disc2.gesdisc.eosdis.nasa.gov/data"

url <- "https://cmr.earthdata.nasa.gov/search/granules"
resp <- httr::GET(url,  query = query, netrc_config)

req <- request(earthdata_api) #%>% 
  #req_dry_run()
req
```

